Think deep for reef resilience

### 2025-08-28 paper/paper_draft.qmd
Initial paper updates to incorporate tracked changes in online draft

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming 
    connectivity was found to influence reef temporal behaviour. Our findings suggest that 
    depth provides critical protection against thermal stress, contrasting current restoration
    practices that primarily target easily accessible shallow sites. These results have 
    important implications for reef management, indicating that depth refugia should be 
    prioritized for further consideration in conservation planning alongside connectivity 
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  custom:
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Climate change is impacting species survival, persistence and ecosystem functions globally, including coral reefs. Coral reefs are vulnerable to increasing magnitude and variation in environmental stressors such as ocean temperature, acidification and nutrient loading [@foster2021; @gattuso2014]. Mass bleaching and mortality events linked to climate change can cause reef-building coral population declines, possibly leading to long-term and large-scale coral-macroalgae regime shifts [@arif2022; @wild2011]. Climate impacts on coral reefs are, and will continue to, decrease the ability of coral reefs to provide ecological functions such as benthic habitat and nutrient cycling [@forster2021; @williams2019] as well as ecosystem services such as food, protection, tourism and Indigenous cultural value [@eddy2021; @stoeckl2021].

@@

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@grimaldi2022]. The GBR faces severe projected coral cover declines under climate change [@bozec2025; @bozec2022]. These coral cover declines will be compounded by ocean acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs) [@cantin2024].

At the individual coral scale, survival during thermal stress events depends on both environmental and biological factors. Environmental factors can either amplify or reduce thermal stress exposure. Depth has been suggested as one crucial factor that can provide a refuge from thermal stress [@álvarez-noriega2018; @baird2018; @smith2014]. Biological factors including coral size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005], and thermal tolerance through acclimation and inheritance [@humanes2022; @lachs2024] all influence survival outcomes. Importantly, corals demonstrate adaptive capacity to develop thermal tolerance that could delay widespread bleaching under moderate emission scenarios [@bairos-novak2021; @humanes2022], representing a crucial mechanism for reef persistence.

@@

The coral demographic representation and associated model implementation are further detailed in the supplementary material. Of important note, however, is that the representation of bleaching mortalities followed Bozec et al., [-@bozec2022] albeit with a scaling coefficient to incorporate the amelioration effect that depth may have, following Baird et al., [@baird2018]. This depth-scaling reduces the proportion of coral colonies impacted by high heat stress as depth increases. The impacts of tropical cyclones and wave activity on reefs were not included in this investigation.

@@

Reefs were clustered into three qualitative categories based on their temporal behaviour indicated by coral cover over time. The three categories were Low, Medium, and High, indicating their relative ability to maintain coral cover and thus their structural resilience. The Complexity-Invariant Distance (CID) between each time series was calculated to identify similarities in temporal behaviour [@batista2014]. The CID applies a correction factor to the Euclidean distance between pairs of time series to obtain a measure of their similarity or dissimilarity. The adopted approach ensures reefs are clustered based on their temporal dynamics as opposed to, for example, naively clustering by their mean coral cover. Clustering similar time series is achieved through the use of k-medoids clustering [cf. @steinmann2020]. Further details on the time series clustering approach are found in the supplementary materials.

>>>>

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef coral cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming 
    connectivity was found to influence reef temporal behaviour. Our findings suggest that 
    depth provides critical protection against thermal stress, contrasting current restoration
    practices that primarily target easily accessible shallow sites. These results have 
    important implications for reef management, indicating that depth refugia should be 
    prioritized for further consideration in conservation planning alongside connectivity 
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  custom:
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Climate change is impacting species survival, persistence and ecosystem functions globally. This is particularly true for coral reefs, which are vulnerable to increasing magnitude and variation in environmental stressors such as ocean temperature, acidification and nutrient loading [@foster2021; @gattuso2014]. Mass bleaching and mortality events driven by climate change can cause reef-building coral population declines, possibly leading to long-term and large-scale coral-macroalgae regime shifts [@arif2022; @wild2011]. Climate impacts on coral reefs are decreasing the ability of coral reefs to provide ecological functions such as benthic habitat and nutrient cycling [@forster2021; @williams2019] as well as ecosystem services such as food, protection, tourism and Indigenous cultural value [@eddy2021; @stoeckl2021].

@@

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@grimaldi2022]. The GBR faces severe projected coral cover declines under climate change [@bozec2025; @bozec2022]. These coral cover declines will be compounded by ocean acidification, tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs) [@cantin2024].

At the individual coral scale, survival during thermal stress events depends on both environmental and biological factors. Environmental factors can either amplify or reduce thermal stress exposure. Depth is a crucial factor that can provide a refuge from thermal stress [@álvarez-noriega2018; @baird2018; @smith2014]. Biological factors including coral size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005], and thermal tolerance through acclimation and inheritance [@humanes2022; @lachs2024] all influence survival outcomes. Importantly, corals demonstrate adaptive capacity to develop thermal tolerance that could delay widespread bleaching under moderate emission scenarios [@bairos-novak2021; @humanes2022], representing a crucial mechanism for reef persistence.

@@

The coral demographic representation and associated model implementation are further detailed in the supplementary material. Of important note, however, is that the representation of bleaching mortalities followed Bozec et al., [-@bozec2022] albeit with a scaling coefficient to incorporate the amelioration effect that depth may have, following Baird et al., [@baird2018]. This depth-scaling reduces the proportion of coral colonies impacted by high heat stress as depth increases. The impacts of tropical cyclones, wave activity, Crown of Thorns Starfish or coral disease were not included in this investigation, in order to focus on assessing reef resilience under possible future warming scenarios.

@@

Reefs were clustered into three qualitative categories based on their temporal behaviour indicated by coral cover over time. The three categories were Low, Medium, and High, indicating their relative ability to maintain coral cover and thus their structural resilience. The Complexity-Invariant Distance (CID) between each time series was calculated to identify similarities in temporal behaviour [@batista2014]. The CID applies a correction factor to the Euclidean distance between pairs of time series to obtain a measure of their similarity. The adopted approach ensures reefs are clustered based on their temporal dynamics as opposed to, for example, naively clustering by their mean coral cover. Clustering similar time series is achieved through the use of k-medoids clustering [cf. @steinmann2020]. Further details on the time series clustering approach are found in the supplementary materials.
